---
title: "Dex Dosage Validation Pilot Results"
output:
  pdf_document: default
  html_document: default
date: "01-03-2020"
---
To validate the dose of Dex to use in my negative feedback / stress series tests, I tested two doses (0.5 mg/kg and 1 mg/kg) on sample sizes of 4 non-breeding birds (each dose). Birds were placed in a bag for 30 minutes as a classic, acute stressor. After removal from the bag, I took a blood sample immediately (0 min), and then administered the vehicle or dex dose intramuscularly. I took blood samples 30, 60, and 90 min after removal from the bag and administration of the DEX. 

```{r setup, include=FALSE}
library(tidyverse)

#functions for graphs 
pd<- position_dodge(0.1) # move them .05 to the left and right

```

```{r data setup, echo = FALSE}
dval<-read.csv("~/Documents/projects/PRL_vs_HPA/2020-02-28_dex_validation_samples_JG_.csv")
dval<-dval %>% separate(sample_label, c("id", "sample_no")) #separate sample label into id and timepoint 
dval$group <- factor(dval$group, levels = c("vehicle", "dex_low", "dex_high"))

#clean up data to remove outliers
dval_trim <- dval %>%
  filter(cort_conc < 45) #removes 2 outliers

#calculate averages, sd, sem, confidence intervals 
dval_stats<-dval_trim %>%
  group_by(group, timepoint) %>%
  summarise(avg_cort = mean(cort_conc), 
            sd_cort = sd(cort_conc), 
            se_cort = sd_cort/sqrt(length(cort_conc)),
            lower_ci = avg_cort - 2*se_cort, 
            upper_ci = avg_cort + 2*se_cort)
```

### Stress series 

```{r individual stress series, echo = FALSE}
#plot overall averages and confidence intervals 
dval_stats %>% 
  filter(group != "dex_low") %>%
ggplot( aes(x=as.factor(timepoint), y=avg_cort, color=group, group = group)) + 
  geom_point(position = pd)+
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), position = pd) +
  geom_line() +
  xlab("Timepoint (min from stressor)")+ 
  ylab("CORT (ng/mL)")+ 
  ggtitle("Average stress series", subtitle = "no significant differences b/w vehicle and high dex dose")+ 
  theme_bw()


#raw stress series by individual
ggplot(dval, aes(x=as.factor(timepoint), y=cort_conc, color=group, group = id)) + 
  geom_point()+
  geom_line() + 
  geom_text(aes(label = id), position = pd)+
  xlab("Timepoint (min from end of 30 min acute stressor)")+ 
  ylab("CORT (ng/mL)")+
  ggtitle("Individual stress series")+ 
  facet_wrap(~group)+
  theme_bw()


```

### Statistics 

There were no significant differences between the vehcile and the low or high dose of DEX (due to the high variablity within the birds). 

```{r stats, echo = FALSE}
#Statistics 
dval_trim$timepoint<-as.factor(dval_trim$timepoint)
dval_trim$timepoint<-relevel(dval_trim$timepoint, "0")
m1<-lm(cort_conc ~ timepoint * group, dval_trim)
summary(m1)
anova(m1)

```
### Possible next steps?

* Increase sample size of Dex validation samples?
* Move forward with 1 mg/kg (the dose used by most Romero studies in the literature)?
* Add a higher dose (2 mg/kg) and re-do validation?

