---
title: "Dex Dosage Validation Pilot Results: Round 2"
output:
  html_document: default
  pdf_document: default
date: "2020-12-02"
---
After an initial pilot study (February 2020) of dexamethasone (DEX) doses did not yield conclusive suppression of CORT release and negative feedback inhibition, I extended the pilot study to run 2 higher doses. 
The previous doses were 0.5 mg/kg and 1 mg/kg. In this pilot study (run October 2020), I used 2 mg/kg and 4 mg/kg. Birds (not actively breeding) were placed in a bag for 30 minutes as a classic acute stressor. After removal from the bag, I immediately took a blood sample (stressed, 0 min of recovery) and then administered the DEX or vehicle (0.09% physiological saline) intramuscularly. I took blood samples 30, 60, and 90 min after removal from the bag and post-administration of the injection. These samples represent the time in recovery (or after DEX/vehicle administration). 

For the sake of fitting all samples on a single RIA run at the metabolomics center, I only ran the 0, 60 and 90 min samples for CORT measurements. (Measured by J.Graham and the UC Davis Metabolomics core)

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(plotrix) #standard error 

#functions for graphs 
pd<- position_dodge(0.1) # move them .05 to the left and right

```

```{r data setup, echo = FALSE}
dex<-read.csv("~/Documents/projects/PRL_vs_HPA/dex_dosage_validation/raw_data/dex_validation_samples_all.csv")

dex <- 
dex %>%
separate(sample_label,into = c("id","sample"), sep = "[.]") %>%
  mutate(group = factor(group, levels = c("vehicle", "0.5 mg/kg", "1 mg/kg", "2 mg/kg", "4 mg/kg")))


#calculate averages, sd, sem, confidence intervals 
dex_stats<-dex %>%
  group_by(group, timepoint) %>%
  summarise(n = n(), 
            avg_cort = mean(cort), 
            se_cort = std.error(cort),
            lower_ci = avg_cort - 1.96*se_cort, 
            upper_ci = avg_cort + 1.96*se_cort)
```

### Stress series 

New pilot data. 

```{r individual stress series, echo = FALSE}
library(RColorBrewer)
my.cols <- brewer.pal(5, "Reds")
my.cols[1]<-"#000000"

#overall averages and CIs
dex_stats %>% 
  filter(group %in% c("vehicle", "2 mg/kg", "4 mg/kg")) %>%
ggplot( aes(x=as.factor(timepoint), y=avg_cort, color=group, group = group)) + 
  geom_point(position = pd)+
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), position = pd) +
  geom_line() +
  xlab("Timepoint (min from stressor)")+ 
  ylab("CORT (ng/mL)")+ 
  scale_color_manual(values = c("#000000", "#FB6A4A","#A50F15")) + 
  annotate("text", x = 3, y = 37, label = "*", size = 7, color = "#A50F15" ) + #significance
  annotate("text", x = 4, y = 37, label = "*", size = 7, color = "#A50F15" ) + 
  ggtitle("Average stress series", subtitle = "Higher DEX doses reduced CORT after 60 min")+ 
  #facet_wrap(~group)+
  theme_bw()

#raw stress series by individual

dex %>% 
  filter(group %in% c("vehicle", "2 mg/kg", "4 mg/kg")) %>%
ggplot(aes(x=as.factor(timepoint), y=cort, color=group, group = id)) + 
  geom_point()+
  geom_line() + 
  geom_text(aes(label = id), position = pd)+
  xlab("Timepoint (min from end of 30 min acute stressor)")+ 
  ylab("CORT (ng/mL)")+
  scale_color_manual(values = c("#000000", "#FB6A4A","#A50F15")) + 
  ggtitle("Individual stress series")+ 
  facet_wrap(~group)+
  theme_bw()


```
### All Pilot Data 

### Statistics 

2 mg/kg dose of DEX successfully lowered plasma CORT levels after 60 and 90 minutes (p < 0.05). 

```{r stats, echo = FALSE}
#Statistics 
dex$timepoint <- as.factor(dex$timepoint)

dex_new <- dex %>% filter(group %in% c("vehicle", "2 mg/kg", "4 mg/kg"))

m1<-lm(cort ~ timepoint * group, dex_new)
summary(m1)
```

Did the DEX successfully lower CORT *at the same timepoint* compared to vehicle-treated birds?

Yes, 2 mg/kg treated birds had significantly lowered CORT after 60 min and 90 min than vehicle-treated birds.
```{r dose fx by timepoint, echo = FALSE}
min60 <- dex_new %>% filter(timepoint == "60") 
pairwise.t.test(min60$cort, min60$group, p.adj = "holm")

min90 <- dex_new %>% filter(timepoint == "90") 
pairwise.t.test(min90$cort, min90$group, p.adj = "holm")
```

Did the DEX succesfully lower CORT compared to stressed (0 min recovery) levels?

Yes, 2 mg/kg and 4 mg/kg doses significantly lowered CORT compared to those doses' average stress-induced CORT levels (0 min after stressor). 

```{r timepoint fx by dose, echo = FALSE}
lowdose <- dex %>% filter(group == "2 mg/kg")
pairwise.t.test(lowdose$cort, lowdose$timepoint)

highdose <- dex %>% filter(group == "4 mg/kg")
pairwise.t.test(highdose$cort, highdose$timepoint)

vehicle <- dex %>% filter(group == "vehicle")
pairwise.t.test(vehicle$cort, vehicle$timepoint)
```