---
title: "Negative feedback to stress across parental care"
author: "Victoria"
date: "7/7/2018"
output: html_document
---
###Flaws with experimental design
1) Range of timepoints ("early inc" = days 1 - 7) collected, rather than a specific timepoint (e.g. day 3)
2) Human observation, rather than camera set up, may bias behavioral results
3) Negative feedback trials may not have gone out far enough... may need to go an hour out (here is where camera may be better)
4) Range of ages of birds (experienced versus unexperienced)

```{r data_setup, echo=FALSE}
library(ggplot2)
library(gridExtra)
library(car)
library(tidyr)
#import data into d_stress 
d_stress<-read.csv((file.choose())) #stress_breeding_CORT_behavior.csv
#trim excess NA's at bottom of csv file
d_stress <- d_stress[ which(d_stress$nest_ID != "NA") , ]
#remove excessive cort values / outliers
d_stress <- d_stress[ which(d_stress$cort2 < 400) , ] 
d_stress <- d_stress[ which(d_stress$cort3 < 90) , ] 

#functions sourced from other scripts
source("~/Documents/statistics/fxns/summarySE.R")
source("~/Documents/statistics/fxns/summarySE_withinsubj.R")
source("~/Documents/statistics/fxns/multiplot.R")

```

```{r wide to long}
#convert data from wide to long, using tidyr
dstress_long <- gather(d_stress, cort, conc, cort1:cort3, factor_key=TRUE)
#gather(dataset, key(made from names of columns),name of new value column, names of source columns to convert, treat key column as a factor)

#renaming cort treatments according to minutes taken
levels(dstress_long$cort)[levels(dstress_long$cort)=="cort1"] <- "0"
levels(dstress_long$cort)[levels(dstress_long$cort)=="cort2"] <- "30"
levels(dstress_long$cort)[levels(dstress_long$cort)=="cort3"] <- "60"

#sorting data frame by bird_id and then cort time point
dstress_long <- dstress_long[order(dstress_long$bird_id, dstress_long$cort), ]
```

```{r stress series}
#run summarySE.R function script first
#summary statistics for dataset (used in line graphs)
stats_period<-summarySE(dstress_long, measurevar="conc", groupvars=c("period","cort"))
stats_sex<-summarySE(dstress_long, measurevar="conc", groupvars=c("sex","cort"))

#within-subjects, repeated measures design, use different summary statistics function
RMstat_period<- summarySEwithin(dstress_long, measurevar="conc", withinvars=c("period","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)
RMstat_sex<- summarySEwithin(dstress_long, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)


#subset data to look at period*sex interactions
#for each period subset, summary statistics by sex
early <- dstress_long[ which(dstress_long$period =='early_inc') , ] 
RMstat_sexE<- summarySEwithin(early, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)

late <- dstress_long[ which(dstress_long$period =='late_inc') , ] 
RMstat_sexL<- summarySEwithin(late, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)

chx <- dstress_long[ which(dstress_long$period =='chicks') , ] 
RMstat_sexC<- summarySEwithin(chx, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)

#GRAPHS
pd<- position_dodge(0.1) # move them .05 to the left and right
#line plot by period
p1<-ggplot(RMstat_period, aes(x=cort, y=conc, colour=period, group=period)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylab("Cort (ng/uL)")+
    xlab("Stress series time point") + theme_bw()

#line plot by sex
p2<-ggplot(RMstat_sex, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylab("Cort (ng/uL)")+
    ylim(0,27.5)+ #normalizes axes so all graphs in multiplot have same y
    xlab("Stress series time point") + ggtitle("All time points") + theme_bw()
#early inc by sex
p3<-ggplot(RMstat_sexE, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylim(0,27.5)+
    ylab("Cort (ng/uL)")+ xlab("Stress series time point")+ ggtitle("Early incubation stress series") + theme_bw()
#late inc by sex
p4<-ggplot(RMstat_sexL, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylim(0,27.5)+
     ylab("Cort (ng/uL)")+ xlab("Stress series time point")+ ggtitle("Late incubation stress series") + theme_bw()
#chicks by sex
p5<-ggplot(RMstat_sexC, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylim(0,27.5)+
    ylab("Cort (ng/uL)")+ xlab("Stress series time point")+ ggtitle("Chick rearing stress series") + theme_bw()
multiplot(p2,p4,p3,p5, cols=2) 

```
###Behavior: Exploratory data analysis
```{r exploratory, echo=FALSE, message=FALSE, warning=FALSE}
label=geom_text(aes(x=900, label="15 minutes\n", y=5), colour="black", angle=90, text=element_text(size=11)) #creates a label for 15 min mark

ggplot(d_stress,aes(x = period ,fill = brooding_status)) + 
    geom_bar(position = "fill")
ggplot(d_stress, aes(x=enter_s, fill=period)) + geom_histogram(alpha=0.5, binwidth = 100) + geom_vline(xintercept=900) + label
ggplot(d_stress, aes(x=brood_s, fill=period)) + geom_histogram(alpha=0.5, binwidth = 100) + geom_vline(xintercept=900) + label

ggplot(d_stress, aes(x=restrained_m, fill=period)) + geom_histogram(alpha=0.5, binwidth = 2) 
ggplot(d_stress, aes(x=recovered_m, fill=period)) + geom_histogram(alpha=0.5, binwidth = 2) 

p1<-ggplot(d_stress, aes(y = b1_s, x = period, fill = period)) + geom_boxplot() 
p2<-ggplot(d_stress, aes(y = b2_s, x = period, fill = period)) + geom_boxplot() 
p3<-ggplot(d_stress, aes(y = b3_s, x = period, fill = period)) + geom_boxplot() 
grid.arrange(p1, p2,p3, nrow = 1)
anova(lm(b1_s~period, data=d_stress))
anova(lm(b2_s~period, data=d_stress))
anova(lm(b3_s~period, data=d_stress))

qqPlot(d_stress$brood_s) #looks like a right skew
d_stress$brood_s.log<-log(d_stress$brood_s) #log transform brood times
qqPlot(d_stress$brood_s.log) #qqPlot for log transformed brood times. Improved normality!

#log transform data in long dataset
dstress_long$brood_s.log<-log(dstress_long$brood_s) 
qqPlot(dstress_long$conc) #cort data has some bad right skew
dstress_long$conc.log<-log(dstress_long$conc)
qqPlot(dstress_long$conc.log)

plot.design(brood_s.log ~ period + brooding_status + sex,
            data = d_stress)

#pair plot of correlational matrix
pairs.panels(d_stress[,c(33,19,27,28,31,32)], #define the columns/variables you want to correlate using their #s
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )

```
```{r}
#does day predict stress-induced cort?
ggplot(d_stress, aes(x=inc_day, y=log(cort1)))+
  geom_point(shape=1) + #hollow circles for points
  geom_smooth(method=lm, se=TRUE) #add regression line without confidence interval shading

ggplot(d_stress, aes(x=inc_day, y=log(cort2)))+
  geom_point(shape=1) + #hollow circles for points
  geom_smooth(method=lm, se=TRUE) #add regression line without confidence interval shading

summary(lm(log(cort1)~ inc_day, data=d_stress)) #trend towards increasing baseline cort with increasing 
summary(lm(log(cort2)~ inc_day, data=d_stress))
summary(lm(log(cort3)~ inc_day, data=d_stress))

```
