---
title: "Negative feedback to stress across parental care"
author: "Victoria"
date: "7/7/2018"
output: html_document
---
###Flaws with experimental design
1) Range of timepoints ("early inc" = days 1 - 7) collected, rather than a specific timepoint (e.g. day 3)
2) Human observation, rather than camera set up, may bias behavioral results
3) Negative feedback trials may not have gone out far enough... may need to go an hour out (here is where camera may be better)
4) Range of ages of birds (experienced versus unexperienced)

```{r data_setup, include=FALSE}
library(ggplot2)
library(cowplot) #multiple graphs in one plot, alternative to multiplot function
library(car) #regressions
library(tidyr) #gather
library(dplyr) #mutate, group_by, summarize
librarY(psych) #pairs.panel (pair plot)
library(plotrix) #standard error calculations
#import data into d_stress 
d_stress<-read.csv("~/Documents/lab/neg feedback parental care/stress_breeding_CORT_behavior.csv") #stress_breeding_CORT_behavior.csv
#trim excess NA's at bottom of csv file
d_stress <- d_stress[ which(d_stress$nest_ID != "NA") , ]
#remove excessive cort values / outliers
d_stress <- d_stress[ which(d_stress$cort2 < 400) , ] 
d_stress <- d_stress[ which(d_stress$cort3 < 90) , ] 

#functions sourced from other scripts
source("~/Documents/statistics/fxns/summarySE.R")
source("~/Documents/statistics/fxns/summarySE_withinsubj.R")
#source("~/Documents/statistics/fxns/multiplot.R") #replaced with cowplot

d_stress$period <- factor(d_stress$period, levels=c("early_inc", "late_inc", "chicks"))
dstress_long$period <- factor(dstress_long$period, levels=c("early_inc", "late_inc", "chicks"))

```

```{r wide to long}
#convert data from wide to long, using tidyr
dstress_long <- gather(d_stress, cort, conc, cort1:cort3, factor_key=TRUE)
#gather(dataset, key(made from names of columns),name of new value column, names of source columns to convert, treat key column as a factor)

#renaming cort treatments according to minutes taken
levels(dstress_long$cort)[levels(dstress_long$cort)=="cort1"] <- "0"
levels(dstress_long$cort)[levels(dstress_long$cort)=="cort2"] <- "30"
levels(dstress_long$cort)[levels(dstress_long$cort)=="cort3"] <- "60"

#sorting data frame by bird_id and then cort time point
dstress_long <- dstress_long[order(dstress_long$bird_id, dstress_long$cort), ]
```

###Behavior: Exploratory data analysis
```{r exploratory, echo=FALSE, message=FALSE, warning=FALSE}
label=geom_text(aes(x=900, label="15 minutes\n", y=5), colour="black", angle=90, text=element_text(size=11)) #creates a label for 15 min mark

#behavior histograms
behav1<-ggplot(d_stress, aes(x=enter_s, fill=period)) + geom_histogram(alpha=0.5, binwidth = 100) + geom_vline(xintercept=900) + label + scale_x_continuous(name="Time to enter (sec)", limits=c(0, 1500))
behav2<-ggplot(d_stress, aes(x=brood_s, fill=period)) + geom_histogram(alpha=0.5, binwidth = 100) + geom_vline(xintercept=900) + label + scale_x_continuous(name="Time to brood (sec)", limits=c(0, 1500))
plot_grid(behav1,behav2, rows=2, align="v")

#brooding behavior by period
brood<- d_stress %>% 
  group_by(period,brooding_status) %>% 
  summarise(count=n()) %>% 
  mutate(perc=count/sum(count)) %>%
  ggplot(aes(x = period, y = perc*100, fill = factor(brooding_status), label= round(perc*100,1))) +
  geom_bar(colour = "black", stat="identity", width = 0.7) +
  geom_text(position = "stack", vjust = 2)+
  labs(x = "Period", y = "Percent", fill = "Response") +
  ggtitle("Percent brooding by period")+
  theme_minimal()
print(brood)

#bleeding time per blood sample, by period
plot1<-ggplot(d_stress, aes(y = b1_s, x = period, fill = period)) + geom_boxplot() 
plot2<-ggplot(d_stress, aes(y = b2_s, x = period, fill = period)) + geom_boxplot() 
plot3<-ggplot(d_stress, aes(y = b3_s, x = period, fill = period)) + geom_boxplot() 
plot_grid(plot1, plot2,plot3, cols=1)
anova(lm(b1_s~period, data=d_stress))
anova(lm(b2_s~period, data=d_stress))
anova(lm(b3_s~period, data=d_stress))

#qqplots for behavior
qqPlot(d_stress$brood_s, main="Brood times (raw)") #looks like a right skew
d_stress$brood_s.log<-log(d_stress$brood_s) #log transform brood times
qqPlot(d_stress$brood_s.log,main="Log transformed brood times") #qqPlot for log transformed brood times. Improved normality!


#log transform data in long dataset (CORT data)
dstress_long$brood_s.log<-log(dstress_long$brood_s) 
qqPlot(dstress_long$conc, main = "Cort concentrations (raw)") #cort data has some bad right skew
dstress_long$conc.log<-log(dstress_long$conc)
qqPlot(dstress_long$conc.log, main="Cort concentrations (log-transformed)")

#design plot
plot.design(brood_s.log ~ period + brooding_status + sex,
            data = d_stress)

#pair plot of correlational matrix
pairs.panels(d_stress[,c(33,19,27,28,31,32)], #define the columns/variables you want to correlate using their #s
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )

```

###Stress series
```{r stress series}
#run summarySE.R function script first
#summary statistics for dataset (used in line graphs)
stats_period<-summarySE(dstress_long, measurevar="conc", groupvars=c("period","cort"))
stats_sex<-summarySE(dstress_long, measurevar="conc", groupvars=c("sex","cort"))

#within-subjects, repeated measures design, use different summary statistics function
RMstat_period<- summarySEwithin(dstress_long, measurevar="conc", withinvars=c("period","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)
RMstat_sex<- summarySEwithin(dstress_long, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)


#subset data to look at period*sex interactions
#for each period subset, summary statistics by sex
early <- dstress_long[ which(dstress_long$period =='early_inc') , ] 
RMstat_sexE<- summarySEwithin(early, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)

late <- dstress_long[ which(dstress_long$period =='late_inc') , ] 
RMstat_sexL<- summarySEwithin(late, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)

chx <- dstress_long[ which(dstress_long$period =='chicks') , ] 
RMstat_sexC<- summarySEwithin(chx, measurevar="conc", withinvars=c("sex","cort"),
                        idvar="bird_id", na.rm=TRUE, conf.interval=.95)

#GRAPHS
pd<- position_dodge(0.1) # move them .05 to the left and right
#line plot by period
p1<-ggplot(RMstat_period, aes(x=cort, y=conc, colour=period, group=period)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylab("Cort (ng/uL)")+
    xlab("Stress series time point") + theme_bw()

#line plot by sex
p2<-ggplot(RMstat_sex, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylab("Cort (ng/uL)")+
    ylim(0,27.5)+ #normalizes axes so all graphs in multiplot have same y
    xlab("Stress series time point") + ggtitle("All time points") + theme_bw()
#early inc by sex
p3<-ggplot(RMstat_sexE, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylim(0,27.5)+
    ylab("Cort (ng/uL)")+ xlab("Stress series time point")+ ggtitle("Early incubation stress series") + theme_bw()
#late inc by sex
p4<-ggplot(RMstat_sexL, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylim(0,27.5)+
     ylab("Cort (ng/uL)")+ xlab("Stress series time point")+ ggtitle("Late incubation stress series") + theme_bw()
#chicks by sex
p5<-ggplot(RMstat_sexC, aes(x=cort, y=conc, colour=sex, group=sex)) + 
    geom_errorbar(aes(ymin=conc-ci, ymax=conc+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylim(0,27.5)+
    ylab("Cort (ng/uL)")+ xlab("Stress series time point")+ ggtitle("Chick rearing stress series") + theme_bw()
plot_grid(p2,p3,p4,p5, align="h")

```
Fascinating! Trend towards birds that did not brood in 30 minutes after a stressor having *lower* stress-induced CORT levels than birds that did brood after the stressor. 
This shows up in the ANOVA as well, with brooding status possibly being significant in the model for cort (p = 0.055, but with no significant interaction with series time point).   
```{r stress series brooding status}
#using dplyr to do what the summarySE scoped-in function does. ISSUE: does not do within-subjects control
stat_brood<- dstress_long %>%
  drop_na(conc)%>%
  group_by(brooding_status,cort) %>%
  summarize(mean=mean(conc), se=std.error(conc), ci=std.error(conc)*1.96)

p6<-ggplot(stat_brood, aes(x=cort, y=mean, colour=brooding_status, group=brooding_status)) + 
    geom_errorbar(aes(ymin=mean-ci, ymax=mean+ci), width=.1, position=pd)  +
    geom_line(position=pd) +
    geom_point(position=pd, size=3) +
    ylab("Cort (ng/uL)")+
    xlab("Stress series time point") + theme_bw()
print(p6)
```
**Note:** Currently, non-brooding birds have brood_s as "NA". Would it change the statistics if brooding birds were listed as NA?
```{r stats}
#cort data model
cort<-lmer(conc.log ~ brooding_status + period + sex + cort + (1|bird_id), data=dstress_long)
Anova(cort)
anova(cort)
#interaction between brooding_status and cort time point?
cort2<-lmer(conc.log ~brooding_status*cort + (1|bird_id), data=dstress_long)
Anova(cort2)


#behavior data model
behav<-lmer(brood_s.log ~ period*sex + (1|bird_id), data=dstress_long) #why is this so much more significant than the base lm? 
Anova(behav)
anova(behav)
anova(lm(brood_s.log~period*sex,data=d_stress))

#behavior data graphs
ggplot(dstress_long, aes(y = brood_s.log, x = period, fill = sex)) + geom_boxplot() #boxplot

# means with confidence intervals
brood.sexXstage<-dstress_long %>%
  drop_na(brood_s.log)%>%
  group_by(period,sex) %>%
  summarize(mean=mean(brood_s.log), se=std.error(brood_s.log), ci=std.error(brood_s.log)*1.96)

p7<-ggplot(brood.sexXstage, aes( x=period, y=mean, colour=sex)) + 
  geom_errorbar(aes(ymin=mean-ci, ymax=mean+ci),width=0.1, position=pd) +
  geom_point(position=pd,size=3) +
  ylab("Log(time to brood nest)") +
  ylim(5,7) +
  theme_minimal()
print(p7)


#does day predict stress-induced cort? Regressions
ggplot(d_stress, aes(x=inc_day, y=log(cort1)))+
  geom_point(shape=1) + #hollow circles for points
  geom_smooth(method=lm, se=TRUE) #add regression line without confidence interval shading

ggplot(d_stress, aes(x=inc_day, y=log(cort2)))+
  geom_point(shape=1) + #hollow circles for points
  geom_smooth(method=lm, se=TRUE) #add regression line without confidence interval shading

summary(lm(log(cort1)~ inc_day, data=d_stress)) #weak trend towards increasing baseline cort with increasing 
summary(lm(log(cort2)~ inc_day, data=d_stress))
summary(lm(log(cort3)~ inc_day, data=d_stress))

```
